---
title: "Activity 3"
author: "Delaney Scheiern"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(lubridate)
```


```{r}
#create a function. The names of the arguments for your function will be in parentheses. Everything in curly brackets will be run each time the function is run.
assert <- function(statement,err.message){
  #if evaluates if a statement is true or false for a single item
  if(statement == FALSE){
    print(err.message)
  }
  
}

#check how the statement works
#evaluate a false statement
assert(1 == 2, "error: unequal values")
```

```{r}
#evaluate a true statement
assert(2 == 2, "error: unequal values")
#set up assert to check if two vectors are the same length
a <- c(1,2,3,4)
b <- c(8,4,5)
assert(length(a) == length(b), "error: unequal length")
```

### Question 1: Describe how the accuracy of the relative humidity measurements vary with temperature and humidity.
*For extreme values of temperature and humidity, the accuracy of relative humidity is much less accurate. For low-mid temperatures, there is a higher range of accurate humidity levels for relative humidity. For mid-high temperatures, there is a narrower range of humidity values that give an accurate relative humidity reading.*

### Question 2: Describe the quality assurance protocols in place for the anemometer. What type of unreliable measurements do they prevent? What are cases where data would be incorrectly marked as unreliable?
*An instantaneous measurement is rejected if it is more than eight times the running average. This prevents unreliable measurements like spikes in wind speed from bugs. This would incorrectly mark data unreliable if there was a huge change in wind speed, like turning on a wind tunnel.*


```{r}
#read in the data file
#skip the first 3 rows since there is additional column info
#specify the the NA is designated differently
datW <- read.csv("data\\bewkes\\bewkes_weather.csv",
          na.strings=c("#N/A"), skip=3, header=FALSE)
#preview data
print(datW[1,])
```


```{r}
#get sensor info from file
# this data table will contain all relevant units
sensorInfo <- read.csv("data\\bewkes\\bewkes_weather.csv",
                na.strings=c("#N/A"), nrows=2)

print(sensorInfo)
```

```{r}
#get column names from sensorInfo table
# and set weather station colnames  to be the same
colnames(datW) <- colnames(sensorInfo)
#preview data
print(datW[1,])
```

### Question 3: What is the difference between skip and nrows in these two read.csv commands? What does header=FALSE do?
*Skip tells the function not to read the specified row into the dataframe. nrows only puts the first number of specified rows into the dataframe. header=FALSE indicates that the first row does not contain the header titles that we want.*

```{r}
#convert to standardized format
#date format is m/d/y
dates <- mdy_hm(datW$timestamp, tz= "America/New_York")
```

```{r}
#calculate day of year
datW$doy <- yday(dates)
#calculate hour in the day
datW$hour <- hour(dates) + (minute(dates)/60)
#calculate decimal day of year
datW$DD <- datW$doy + (datW$hour/24)
#quick preview of new date calcualtions
datW[1,]
```

```{r}
#see how many values have missing data for each sensor observation
#air temperature
length(which(is.na(datW$air.temperature)))
```

```{r}
#wind speed
length(which(is.na(datW$wind.speed)))
```

```{r}
#precipitation
length(which(is.na(datW$precipitation)))
```

```{r}
#soil temperature
length(which(is.na(datW$soil.moisture)))
```

```{r}
#soil moisture
length(which(is.na(datW$soil.temp)))
```

```{r}
#make a plot with filled in points (using pch)
#line lines
plot(datW$DD, datW$soil.moisture, pch=19, type="b", xlab = "Day of Year",
     ylab="Soil moisture (cm3 water per cm3 soil)")
```

```{r}
#make a plot with filled in points (using pch)
#line lines
plot(datW$DD, datW$air.temperature, pch=19, type="b", xlab = "Day of Year",
     ylab="Air temperature (degrees C)")
```

```{r, include=FALSE}
#I'm going to make a new column to work with that indicates that I am conducting QAQC
#because overwriting values should be done cautiously and can lead to confusing issues.
#It can be particularly confusing when you are just learning R.
#Here I'm using the ifelse function
#the first argument is a logical statement to be evaluated as true or false on a vector
#the second argument is the value that my air.tempQ1 column will be given if the statement
#is true. The last value is the value that will be given to air.tempQ1 if the statement is false.
#In this case it is just given the air temperature value
datW$air.tempQ1 <- ifelse(datW$air.temperature < 0, NA, datW$air.temperature)
```

```{r}
#check the values at the extreme range of the data
#and throughout the percentiles
quantile(datW$air.tempQ1)
```

```{r}
#look at days with really low air temperature
datW[datW$air.tempQ1 < 8,] 
```
*According to https://www.wunderground.com/history/daily/us/ny/hamilton/KNYHAMIL11/date/2019-11-4, the low on 6/26/2018 near Hamilton was 44 degrees F. Our data shows that the low temperature on this day was 7.0 degrees C, or 44.6 degrees F, which is very close.*


```{r}
 #look at days with really high air temperature
datW[datW$air.tempQ1 > 33,]  
```
*According to https://www.wunderground.com/history/daily/us/ny/hamilton/KNYHAMIL11/date/2019-11-4, the high on 7/1/2018 near Hamilton was 95 degrees F. Our data shows that the high temperature on this day was 34.4 degrees C, or 93.9 degrees F, which is very close.*

### Question 4: Are the extreme high and low values in this dataset reliably measured by the sensor? Explain your answer.
*The extreme high and low values in this dataset are reliably measured by the sensor. The air temperature sensor has a range of -50 to 60 degrees C, which is a very inclusive and large range for air temperature in this area.*


```{r}
#plot precipitation and lightning strikes on the same plot
#normalize lighting strikes to match precipitation
lightscale <- (max(datW$precipitation)/max(datW$lightning.acvitivy)) * datW$lightning.acvitivy
#make the plot with precipitation and lightning activity marked
#make it empty to start and add in features
plot(datW$DD , datW$precipitation, xlab = "Day of Year", ylab = "Precipitation & lightning",
        type="n")
#plot precipitation points only when there is precipitation 
#make the points semi-transparent
points(datW$DD[datW$precipitation > 0], datW$precipitation[datW$precipitation > 0],
        col= rgb(95/255,158/255,160/255,.5), pch=15)        
        
#plot lightning points only when there is lightning     
points(datW$DD[lightscale > 0], lightscale[lightscale > 0],
        col= "tomato3", pch=19)
```

### Question 5: The variable lightscale used to graph lightning values on the same plot as precipitation is not in the datW dataframe. Explain why you can still use it to subset values in datW. Provide evidence for your answer by creating a test that uses your assert function from part 1.

*lightscale > 0 returns an array of booleans that represent for each index if there was lightning or not. The True indexes are used to subset datW. Since lightscale has the same length and order of observations as datW, we can use it to subset datW.*
```{r}
assert(length(lightscale > 0) == length(datW$timestamp), err.message = "Unequal lengths")
assert(length(datW$DD[lightscale > 0]) == sum(lightscale > 0), err.message = "Unequal number of Trues")
```


```{r}
# filter out storms in wind and air temperature measurements
# filter all values with lightning that coincides with rainfall greater than 2mm or only rainfall over 5 mm.    
# create a new air temp column
datW$air.tempQ2 <- ifelse(datW$precipitation  >= 2 & datW$lightning.acvitivy >0, NA, ifelse(datW$precipitation > 5, NA, datW$air.tempQ1))
```

### Question 6: Repeat the code above to remove suspect measurements from wind speed measurements. Create a test using assert to verify that this filtered the data as expected. Describe your outcome. Include a plot with both lines and points of windspeed with the new data.
```{r}
#filter out suspect measurements from wind speed.
quantile(datW$wind.speed)
datW$wind.speed2 <- ifelse(datW$wind.speed < 0.1, NA, ifelse(datW$wind.speed > 1.5, NA, datW$wind.speed))
#assert(datW$wind.speed==datW$wind.speed2, err.message = "Columns are not the same anymore")
plot(datW$DD, datW$wind.speed2, pch=19, type="b", xlab = "Day of Year",
     ylab="Wind Speed")
```


### Question 7: Check that the soil temperature and moisture measurements are reliable in the days leading up to the sensor outage. Explain your reasoning and show all code. Keep in mind that precipitation and air temperature can be used to help provide context.
```{r}
range(datW$soil.temp, na.rm = TRUE)
range(datW$soil.moisture, na.rm = TRUE)
head(datW[datW$soil.moisture > 0.26,])
head(datW[datW$precipitation > 5,])
```
*The soil temperature and moisture measurements before the sensor outage are accurate. The ranges of values are well within the measurable ranges. The values for soil moisture spike with precipitation, as expected.*

### Question 8: The researchers requested a table with the average air temperature, wind speed, soil moisture, and soil temperature for their study period. They would also like the total precipitation. Indicate how many observations went into these calculations and the time period of measurement. Report your findings with the correct number of decimal places that is within the sensor error.
```{r}
library(tidyverse)
mySumm <- datW %>% summarize(avg.airtemp = round(mean(air.temperature),1),
                             avg.windspeed = round(mean(wind.speed),2),
                             avg.soilmoisture = round(mean(soil.moisture, na.rm = TRUE),2),
                             avg.soiltemp = round(mean(soil.temp, na.rm = TRUE),2),
                             total.precip = round(sum(precipitation, na.rm = TRUE),3))
mySumm
range(datW$DD)
length(datW$DD)
```
*2118 observations went into these calculations. The date range was from the 163rd day of the year to the 207th day of the year.*

### Question 9: Make four plots of soil moisture, air temperature, soil temperature, and precipitation throughout all observations in the study period. Use the same x axis range for each plot. In a few sentences, briefly describe trends in the data.
```{r}
par(mfrow=c(2,2))
plot(datW$DD, datW$soil.moisture, pch=19, type="b", xlab = "Day of Year",
     ylab="Soil Moisture")
plot(datW$DD, datW$soil.temp, pch=19, type="b", xlab = "Day of Year",
     ylab="Soil Temperature")
plot(datW$DD, datW$air.temperature, pch=19, type="b", xlab = "Day of Year",
     ylab="Air Temperature")
plot(datW$DD, datW$precipitation, pch=19, type="b", xlab = "Day of Year",
     ylab="Precipitation")

```
*Soil moisture seems to peak on days with lots of precipitation, then slowly decrease. Soil temperature and air temperature seem to follow the exact same pattern, just at a different scale.*

### Question 10: Copy the URL of your Rscript in your GitHub repository.
https://github.com/dscheiern/GEOG331/blob/master/Activity%203.Rmd
