---
title: "Activity 6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages
library(raster)
library(sp)
library(rgdal)
library(rgeos)
library(plyr)
```


```{r}
#read in shapefiles
#readOGR in rgdal does this
g1966 <- readOGR("data\\GNPglaciers\\GNPglaciers_1966.shp", stringsAsFactors = T)

g1998 <- readOGR("data\\GNPglaciers\\GNPglaciers_1998.shp", stringsAsFactors = T)

g2005 <- readOGR("data\\GNPglaciers\\GNPglaciers_2005.shp", stringsAsFactors = T)

g2015 <- readOGR("data\\GNPglaciers\\GNPglaciers_2015.shp", stringsAsFactors = T)
```

```{r}
head(g2015@data)
```


```{r}
g2015@polygons[[1]]
```

```{r}
g1966@proj4string
```

### Question 1: Look up the projection and datum. What does zone mean? Describe the projected coordinate system and the properties that it maintains. What spatial scale is this projected coordinate system meant for?
*Answer: The projection is UTM (Universal Transverse Mercator). This is a horizontal position representation and treats the earth as a perfect ellipsoid. It divides earth into 60 zones. Zone 12 is one of these zones. NAD 83 is the most current datum that provides spatial reference for Canada and the US using latitude, longitude, and some height info.*

```{r}
spplot(g1966, "GLACNAME")
```

```{r}
#fix glacier name so that it is consistent with the entire time period
g2015@data$GLACNAME <- ifelse(g2015@data$GLACNAME == "North Swiftcurrent Glacier", "N. Swiftcurrent Glacier",
           ifelse(g2015@data$GLACNAME ==  "Miche Wabun", 
                        "Miche Wabun Glacier",
                    as.character(g2015@data$GLACNAME)))
```

```{r}
#read in rgb imagery from landsat
redL <- raster("data\\glacier_09_05_14\\l08_red.tif")
greenL <- raster("data\\glacier_09_05_14\\l08_green.tif")
blueL <- raster("data\\glacier_09_05_14\\l08_blue.tif")
```


```{r}
redL@crs
```

```{r}
#make a brick that stacks all layers
rgbL <- brick(redL, greenL, blueL)
```

```{r}
#plot with color
#show axes for reference
#add contrast to the imagery to see it better
par(mai=c(1,1,1,1))
plotRGB(rgbL, stretch="lin", axes=TRUE)
#add polygons to plot
plot(g1966, col="tan3", border=NA, add=TRUE)
plot(g1998, col="royalblue3", add=TRUE, border=NA)
plot(g2005, col="darkgoldenrod4", add=TRUE, border=NA)
plot(g2015, col="tomato3", add=TRUE, border=NA)
```

```{r}
plotRGB(rgbL, ext=c(289995,310000,5371253,5400000), stretch="lin")
plot(g1966, col="palegreen2", border=NA, add=TRUE)
plot(g1998, col="royalblue3", add=TRUE, border=NA)
plot(g2005, col="darkgoldenrod4", add=TRUE, border=NA)
plot(g2015, col="tomato3", add=TRUE, border=NA)
```

```{r}
#set up years to read in
ndviYear <- seq(2003,2016)

#read all files into a list
NDVIraster <- list() 
for(i in 1:length(ndviYear)){
    NDVIraster[[i]] <- raster(paste0("data\\NDVI\\NDVI_",ndviYear[i],".tif"))
}
```

```{r}
NDVIraster[[1]]@crs
```

### Question 2: Why do you want to work with glacier data in an equal area projection rather than the UTM projection?
*Answer: *


```{r}
plot(NDVIraster[[1]])
```

### Question 3: What happens if you try to plot NDVI with the 1966 glacier polygons? Make a plot of the 2003 NDVI data side by side with the 1966 glacier extent. Explain why you can't put both data files on the same map.

```{r}
par(mai=c(1,1,1,1), mfrow = c(1,2))
plot(NDVIraster[[1]])

plotRGB(rgbL, stretch="lin", axes=TRUE)
#add polygons to plot
plot(g1966, col="tan3", border=NA, add=TRUE)
```
The data files can't be put on the same map because they are different projections. You can see that the x and y coordinates for the plots are completely different.



```{r}
#reproject the glaciers
#use the NDVI projection
#spTransform(file to project, new coordinate system)
g1966p <- spTransform(g1966,NDVIraster[[1]]@crs)
g1998p <- spTransform(g1998,NDVIraster[[1]]@crs)
g2005p <- spTransform(g2005,NDVIraster[[1]]@crs)
g2015p <- spTransform(g2015,NDVIraster[[1]]@crs)
```

### Question 4: Make a map with both the maximum NDVI and the glaciers in 2015. Don't show the axis labels with the x and y coordinate system values. Make the 2015 glacier ploygon with no fill color and a black border. What are the patterns in NDVI in the map and around glaciers.

```{r}
plotRGB(rgbL, stretch="lin", axes=FALSE)
#add polygons to plot
plot(g2015, col="transparent", border="black", add=TRUE)
plot(NDVIraster[[13]], add=TRUE)
```

```{r}
#calculate area for all polygons
#add directly into data table for each shapefile
g1966p@data$a1966m.sq <- area(g1966p)
g1998p@data$a1998m.sq <- area(g1998p)
g2005p@data$a2005m.sq <- area(g2005p)
g2015p@data$a2015m.sq <- area(g2015p)
```

```{r}
gAllp1 <- join(g1966p@data,g1998p@data, by="GLACNAME", type="full")
gAllp2 <- join(gAllp1,g2005p@data, by="GLACNAME", type="full")
gAll <- join(gAllp2,g2015p@data, by="GLACNAME", type="full")
```

```{r}
plot(c(1966,1998,2005,2015), 
        c(gAll$a1966m.sq[1],gAll$a1998m.sq[1], gAll$a2005m.sq[1],gAll$a2015m.sq[1]),
        type="b", 
        pch=19, col=rgb(0.5,0.5,0.5,0.5), xlim= c(1965,2016),
        ylim=c(0,2000000),
        ylab="Area of glacier (meters squared)",
        xlab="Year")
        
for(i in 2:39){
points(c(1966,1998,2005,2015), 
        c(gAll$a1966m.sq[i],gAll$a1998m.sq[i], gAll$a2005m.sq[i],gAll$a2015m.sq[i]),
        type="b", 
        pch=19, col=rgb(0.5,0.5,0.5,0.5))
}
```

### Question 5: Calculate the % change in area between 1966 and 2015. Make a spplot of the glaciers in 2015 showing the % change that each glacier has experienced.

```{r}
(percChange <- (sum(g2015p@data$a2015m.sq) - sum(g1966p@data$a1966m.sq))/sum(g2015p@data$a2015m.sq))
```





