---
title: "Activity 6"
author: Delaney Scheiern
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up Spatial Packages

```{r, warning=FALSE}
# load packages
library(raster)
library(sp)
library(rgdal)
library(rgeos)
library(plyr)
```

## Read in spatial data
```{r}
#read in shapefiles
#readOGR in rgdal does this
g1966 <- readOGR("data\\GNPglaciers\\GNPglaciers_1966.shp", stringsAsFactors = T)
g1998 <- readOGR("data\\GNPglaciers\\GNPglaciers_1998.shp", stringsAsFactors = T)
g2005 <- readOGR("data\\GNPglaciers\\GNPglaciers_2005.shp", stringsAsFactors = T)
g2015 <- readOGR("data\\GNPglaciers\\GNPglaciers_2015.shp", stringsAsFactors = T)
```

## Explore data
```{r}
head(g2015@data)
```

```{r}
#projection
g1966@proj4string
```

### Question 1: Look up the projection and datum. What does zone mean? Describe the projected coordinate system and the properties that it maintains. What spatial scale is this projected coordinate system meant for?

*Answer: The projection is UTM (Universal Transverse Mercator). This is a horizontal position representation and treats the earth as a perfect ellipsoid. It divides earth into 60 zones. Zone 12 is one of these zones. NAD 83 is the most current datum that provides spatial reference for Canada and the US using latitude, longitude, and some height info. It is meant for a large or medium scale.*

## Prepare data for visualization and analysis
```{r}
#fix glacier name so that it is consistent with the entire time period
g2015@data$GLACNAME <- ifelse(g2015@data$GLACNAME == "North Swiftcurrent Glacier", "N. Swiftcurrent Glacier",
           ifelse(g2015@data$GLACNAME ==  "Miche Wabun", 
                        "Miche Wabun Glacier",
                    as.character(g2015@data$GLACNAME)))
```

```{r}
#read in rgb imagery from landsat
redL <- raster("data\\glacier_09_05_14\\l08_red.tif")
greenL <- raster("data\\glacier_09_05_14\\l08_green.tif")
blueL <- raster("data\\glacier_09_05_14\\l08_blue.tif")
```


```{r}
#check coordinate system
redL@crs
```

```{r}
#make a brick that stacks all layers
rgbL <- brick(redL, greenL, blueL)
```

## Practice visualizing glacier data
```{r}
# plot same as above with different extent
plotRGB(rgbL, ext=c(289995,310000,5371253,5400000), stretch="lin")
plot(g1966, col="palegreen2", border=NA, add=TRUE)
plot(g1998, col="royalblue3", add=TRUE, border=NA)
plot(g2005, col="darkgoldenrod4", add=TRUE, border=NA)
plot(g2015, col="tomato3", add=TRUE, border=NA)
```

## Read in raster data
```{r}
#set up years to read in
ndviYear <- seq(2003,2016)

#read all files into a list
NDVIraster <- list() 
for(i in 1:length(ndviYear)){
    NDVIraster[[i]] <- raster(paste0("data\\NDVI\\NDVI_",ndviYear[i],".tif"))
}
```

```{r}
# check projection
NDVIraster[[1]]@crs
```

### Question 2: Why do you want to work with glacier data in an equal area projection rather than the UTM projection?

*Answer: UTM is a conformal projection, so it preserves shape rather than area. An equal area projection is important for us working with glacier data because we are focused on analyzing change in area over time, which will be inaccurate with UTM.*


### Question 3: What happens if you try to plot NDVI with the 1966 glacier polygons? Make a plot of the 2003 NDVI data side by side with the 1966 glacier extent. Explain why you can't put both data files on the same map.

```{r}
par(mai=c(1,1,1,1), mfrow = c(1,2))
plot(NDVIraster[[1]])
title("NDVI data")

plotRGB(rgbL, stretch="lin", axes=TRUE)
#add polygons to plot
plot(g1966, col="royalblue", border=NA, add=TRUE)
title("1966 Glacier Data")
```

The glaciers will not appear on top of the NDVI data. The data files can't be put on the same map because they are different projections. You can see that the x and y coordinates for the plots are completely different.

## Reproject the glaciers to help with visualization and analysis
```{r}
#reproject the glaciers
#use the NDVI projection
#spTransform(file to project, new coordinate system)
g1966p <- spTransform(g1966,NDVIraster[[1]]@crs)
g1998p <- spTransform(g1998,NDVIraster[[1]]@crs)
g2005p <- spTransform(g2005,NDVIraster[[1]]@crs)
g2015p <- spTransform(g2015,NDVIraster[[1]]@crs)
```

### Question 4: Make a map with both the maximum NDVI and the glaciers in 2015. Don't show the axis labels with the x and y coordinate system values. Make the 2015 glacier ploygon with no fill color and a black border. What are the patterns in NDVI in the map and around glaciers.

```{r}
#add polygons to plot
plot(NDVIraster[[13]], ext=g2015p, axes=FALSE)
plot(g2015p, col="transparent", border="black", add=TRUE)

```

*Answer: The glaciers are in areas with the lowest NDVI, and areas around glaciers have gradually larger levels of NDVI.*

## Analyze glaciers
```{r}
#calculate area for all polygons
#add directly into data table for each shapefile
g1966p@data$a1966m.sq <- area(g1966p)
g1998p@data$a1998m.sq <- area(g1998p)
g2005p@data$a2005m.sq <- area(g2005p)
g2015p@data$a2015m.sq <- area(g2015p)
```

```{r}
# join data into table
gAllp1 <- join(g1966p@data,g1998p@data, by="GLACNAME", type="full")
gAllp2 <- join(gAllp1,g2005p@data, by="GLACNAME", type="full")
gAll <- join(gAllp2,g2015p@data, by="GLACNAME", type="full")
```

### Question 5: Calculate the % change in area between 1966 and 2015. Make a spplot of the glaciers in 2015 showing the % change that each glacier has experienced.

```{r, warning=FALSE}
gAll$percChange <- (gAll$a2015m.sq - gAll$a1966m.sq)/gAll$a1966m.sq

# put inside raster
g2015p@data$percChange <- gAll$percChange
# make color percChange
spplot(g2015p, "percChange")
```

```{r}
# find glacial difference over years
diffPoly <- gDifference(g1966p, g2015p, checkValidity = 2L)
```

```{r}
#plot with NDVI
plot(NDVIraster[[13]], axes=FALSE, box=FALSE)
plot(diffPoly,col="black", border=NA,add=TRUE)
```

### Question 6: Find the glacier with the largest \% loss. Make a map that best displays the glacial extent for all years for that glacier with the highest \% loss. Include background imagery. You can use the original data for the map to match with the imagery. Add a map title that includes the \% loss and glacier name. The subset function will be helpful here.

```{r}
# find glacier with largest percent loss
lossIndex <- which.min(gAll$percChange)
lossVal <- min(gAll$percChange)
boul <- gAll[lossIndex,]
# subset the glacial data for each year to be that glacier
gmax1966 <- subset(g1966, g1966@data$GLACNAME == boul$GLACNAME)
gmax1998 <- subset(g1998, g1998@data$GLACNAME == boul$GLACNAME)
gmax2005 <- subset(g2005, g2005@data$GLACNAME == boul$GLACNAME)
gmax2015 <- subset(g2015, g2015@data$GLACNAME == boul$GLACNAME)


plotRGB(rgbL, ext=gmax1966@bbox, stretch="lin")
# add title while axes are off
legend("top", legend=NA, title=paste(round(-100*lossVal, 2), "% Lost by", boul$GLACNAME), bty = "n")

#add polygons to plot
plot(gmax1966, col="transparent", border="black", add=TRUE)
plot(gmax1998, col="transparent", border="tan", add=TRUE)
plot(gmax2005, col="transparent", border="royalblue", add=TRUE)
plot(gmax2015, col="transparent", border="red", add=TRUE)
legend("topright",legend=c("1966","1998","2005","2015"), lty=1, col=c("black","tan","royalblue","red"), cex=0.8)

```

## Analyze vegetation growth
```{r}
#extract NDVI values
NDVIdiff <- list()
meanDiff <- numeric(0)
#loop through all NDVI years
for(i in 1:length(ndviYear)){
  #get raster values in the difference polygon
    NDVIdiff[[i]] <- extract(NDVIraster[[i]],diffPoly)[[1]]
    #calculate the mean of the NDVI values
    meanDiff[i] <- mean(NDVIdiff[[i]], na.rm=TRUE)
}
```

```{r}
plot(ndviYear, meanDiff, type="b",
    xlab= "Year",
    ylab="Average NDVI (unitless)",
    pch=19)
```

```{r}
#designate that NDVIraster list is a stack
NDVIstack <- stack(NDVIraster)
#set up lm function to apply to every cell
#where x is the value of a cell
#need to first skip NA values (like lakes)
#if NA is missing in first raster, it is missing in all
#so we can tell R to assign an NA rather than fitting the function
timeT <- ndviYear
fun <- function(x) {
    if(is.na(x[1])){
        NA}else{
        #fit a regression and extract a slope
            lm(x ~ timeT)$coefficients[2]}}
#apply the slope function to the rasters
NDVIfit <- calc(NDVIstack,fun)
#plot the change in NDVI
plot(NDVIfit, axes=FALSE)
```

### Question 7: What are the patterns in maximum NDVI change across the park? Although this type of quick assessment is helpful, what additional information for the regression would you need to more accurately assess this question?
*Answer: Much of the change across the park is minuscule. It seems like the greatest increase in vegetation is occurring relatively far from the glaciers, which I believe is where there is runoff from the glaciers. Additional information that would be helpful is the slope of the terrain.*


```{r}
#buffer glaciers
glacier500m <- gBuffer(g1966p,#data to buffer
                    byid=TRUE,#keeps original shape id 
                    width=500)#width in coordinate system units
```

```{r}
#convert to a raster
buffRaster <- rasterize(glacier500m,#vector to convert to raster
                    NDVIraster[[1]], #raster to match cells and extent
                    field=glacier500m@data$GLACNAME, #field to convert to raster data
                    background=0)#background value for missing data
plot(buffRaster)
```

```{r}
#rasterize glaciers
glacRaster <- rasterize(g1966p, NDVIraster[[1]], field=g1966p@data$GLACNAME, background=0)
#subtract buffer from original glacier
glacZones <- buffRaster - glacRaster
plot(glacZones)
```

### Question 8: Describe what glacZones looks like. How did the raster math accomplish removing the glaciers from the zones. What are the similarities and differences between this raster operation and gDifference?
*Answer: glacZones looks like the outline of the glaciers from 1966. Raster data is stored as pixels with specific values. Since buffRaster and glacZones were both created using the first NDVIraster to match cells and extent, they are directly comparable. When they are are subtracted, the glacier pixel values are canceled out and they disappear. Similarly, gDifference returns the regions of the first geometries that are not in the second geometries.*

```{r}
# find change in area around glaciers
meanChange <- zonal(NDVIfit, #NDVI function to summarize
                glacZones,#raster with zones
                "mean")#function to apply
head(meanChange)
```

### Question 9: Add the mean change in NDVI per year into the 2015 glacier polygons. Make a map where the mean change in vegetation is color coded within the 2015 glacier polygons. Does there seem to be any patterns?
```{r, warning=FALSE}
g2015p@data$meanChange <- meanChange[2:40,"mean"]

spplot(g2015p, "meanChange")
```

*Answer: There doesn't seem to be much of a pattern. Most values seem to be very small or negative, while there are a couple glaciers with positive change. This aligns with the previous observation that most of the increase in NDVI was away from the glaciers.*


### Question 10: Do you think vegetation is changing as glaciers recede and why? Are the values of yearly change substantial? Hint: consider the range of NDVI and also consider the magnitude of change over a 10 or 15 year period.
*Answer: I think it is difficult to conclude that vegetation is changing as glaciers recede over this small of a time period. The mean change in most areas is below the level that was described as substantial in Anderson's paper. However, although small, there is clearly a pattern of vegetation growth from runoff of melting glaciers.*


### Question 11: What is the average maximum NDVI across all years within the Glacier National Park? Is there a pattern between glacier size and NDVI within 500m? Make a map that shows both the recent glacier extent color coded with the maximum NDVI range, and the raster of average maximum NDVI.

```{r}
# take average value across all years
NDVImeans <- calc(NDVIstack,fun=mean)
plot(NDVImeans)

# Scatterplot of glacier size and NDVI within 500
meanVals <- zonal(NDVImeans, #NDVI function to summarize
                buffRaster,#raster with zones
                "mean")#function to apply
plot(g1966p@data$a1966m.sq, meanVals[2:40,"mean"],
    xlab= "Glacier Size",
    ylab="Mean NDVI of Buffer",
    pch=19)
```

*Answer: Yes, there is a pattern between glacier size and NDVI within 500m. As glacier size increases, NDVI decreases exponentially.*

```{r}
means <- meanVals[2:40, "mean"]
g2015p@data$colors <- ifelse(means < 0.2, "red",
                             ifelse(means < 0.4, "purple", "blue"))

plot(NDVImeans, ext=g2015p, axes=FALSE)
plot(g2015p, border=g2015p@data$colors, add=TRUE)
legend("topright",
       legend = c("NDVI < 0.2", "0.2 <= NDVI < 0.4", "NDVI >=0.4"), cex = 0.4, fill=c("red", "purple", "blue"))

```

### Question 12: Using the Carlson and Anderson paper, what kinds of data would you need to validate what is happening with vegetation as glaciers recede? What might be the next steps in making a more accurate analysis of vegetation change for research?
*Answer: To help us validate what is happening with vegetation as glaciers recede, we need elevation data. We also need to know where there is persistent snow coverage in addition to glaciers. Next steps would be to include the terrain slopes and account for the impact of snow cover on vegetation growth.*

### Question 13: Copy and paste the link to your GitHub code.
https://github.com/dscheiern/GEOG331/blob/master/Activity6.Rmd





